Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest

### 2020-11-06 manuscript.rmd
1. Re-arranged Tables and Figures sequence
1. In reference to Psi using more/less negative rather than above/below 2. mortality results include explanation for 1990-1995 census 3. figure captions edited

Section  | Word Count  
------------- | -------------  
Summary | 197  
Main Text  | 6532  
Introduction | 828  
Methods | 3278  
Results | 855  
Discussion | 1571  
Keywords | 6

\
Item | Count  
------------- | -------------  
Figures in color | 9  
Tables | 2  
Supplementary Material |  1 file

@@

To obtain the daily dynamics of soil water potential in the whole-soil column, we locally parameterized and calibrated a 1-D hydrological water balance model. We then developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig. \@ref(fig:schematic)). We define ERD as the depth at which soil water potential best explained species-specific growth dynamics. The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf vulnerability curves by using existing data for eight species, while developing trait-based proxies for the rest (based on data for a total of 21 species). We validated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as independent observations. We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets for six to eight species that overlapped with our assessment of ERD. We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity. Finally, we analysed species-specific large tree mortality dynamics over 35 years and estimated the time a species’ spent below its critical hydraulic threshold to test whether ERD explained risk of drought-induced mortality.

@@

We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up. Extreme hydrological droughts were identified by depth $z$ as days that fell below the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.

@@

\begin{equation}
\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big(z\,\Big|\, \textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|z,h}| r_{s|z,h} \ge 0.71\Big)\Big)
(\#eq:erd)
\end{equation}
where $L$ is the likelihood of $z$ given the observed growth time series, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m. We removed finely resolved soil layers below 0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m as sufficient resolution for Model validation.

@@

\
We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ with ($n$ = `r nrow(mrate.depth.mean)`) and without deciduous species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`). To assess whether deeper ERD species escaped the experience of significant hydrological drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~. $\mathrm{\Psi}$~crit~ was defined as $\mathrm{\Psi}$~50,leaf~.

@@

Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species mortality rates significantly decreased with ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig. \@ref(fig:erd-mort-census-evg)). When deciduous species were included, ERD explained lower but still significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig. \@ref(fig:erd-mort-census)). Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme hydrological droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig. \@ref(fig:swp-droughts)). $\mathrm{\Psi}_{\textrm{soil},z}$ over 1985-1990 is not used for analyses, as that was the ELM-FATES spin-up period, while reliable climate forcing data was unavailable for the prior years. However, the difference between precipitation and ET shows that 1985-1990 had three extreme deficit water years, apart from the 1983 El-Niño leading to it [@Condit.2017]. Analysis of a species' time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig. \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)). Species with ERD deeper than 2 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa).

@@

\
Our result--of lower mortality in deep-rooted trees--contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a south-asian seasonally tropical dry forest had higher mortality in a rare, prolonged drought. The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019 and @Ivanov:2012). In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m never went below $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa) for all the dry seasons and droughts. Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)`mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI are major factors in the different mortality responses. At BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths), \@ref(fig:swp-droughts)), whereas in CT2018 they were not. The contrast between the two studies highlights the combined role of seasonal precipitation input and site-specific hydrology in modulating deep ERD mortality-risk.

@@

\
Figure \@ref(fig:hydro-calib): ELM-FATES calibration and evaluation. Observations (red lines and points) versus simulations (gray lines) from 100 best-fit ensemble-member runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical TDR probe (0-0.15 m; B); monthly stream discharge (Discharge, C) and monthly evapotranspiration from eddy flux tower (ET, D). Red points in (A) show average, manual plot-wide observations of VWC with 95% CI (error bars). All observations (red lines and points) were in-sample, except for VWC data from the horizontal probes (red lines in A) which were out-of-sample. Values in inset are average RMSE across the 100 best-fit simulations. For panel a, values at the top are for manual, plot-wide VWC and those at the bottom are for VWC from TDR probes. VWC is in units of cm^3^ cm^-3^, while ET and Discharge in mm month^-1^.

\
Figure \@ref(fig:swp-droughts): Occurrence of extreme hydrological droughts by census intervals (horizontal panels) at three representative depths ($z$): 0.1, 0.6 and 1 m (vertical panels). Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\mathrm{\Psi}_{\textrm{soil},z}$) for a given DOY and depth $z$ are shown (same across census intervals). $\mathrm{\Psi}_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if, at least one day of that year $\mathrm{\Psi}_{\textrm{soil},z}$ declined below the 5^th^ percentile for that day of year (DOY).

\
Figure \@ref(fig:ab-fit): Parameters $A$ (panel a) & $B$ (panel b) for the leaf vulnerability curves that were fitted to observed data on $K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~ for `r length(sp.exp.param$sp)` species (Eq. \@ref(eq:vc-exp)) vs. those that were predicted from a set of models based on trait-proxies. (Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod)). Goodness-of-fit ($R$^2^) and $p$-values are given in inset.

\
Figure \@ref(fig:erd-iso): Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry-season stable isotopic concentration $\mathrm{\delta}$^2^H~xylem~ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island. Data from [@Meinzer:1999].

\
Figure \@ref(fig:sp-erd): Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r  length(erd.sp)` large, canopy species of Barro Colorado Island. Species ERD is a median of the species' ERDs (depths) obtained from the 100 best-fit hydrological realizations, which are vertically resolved into exponentially increasing depths. Hence, the non-continuous nature of the ERD axis.

\
Figure \@ref(fig:erd-hyd): Pairwise correlations among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama). Spearman's $r$ with red shades for positive $r$ values and blue shades for negative values. $R$^2^ and $p$-values for linear model fits are given in panel insets. See Fig. S\@ref(fig:erd-hyd-full) for the full correlation matrix.

\
Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE in % yr^-1^) versus Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015). $R$^2^ and $p$-values for linear model fits are given in panel insets. Model fits (blue lines) with SE (gray area) are shown for significant fits at $\alpha$ = 0.1.

\
Figure \@ref(fig:erd-time): Modeled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--% days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was below $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~50,leaf~, over each of the five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy tree species on Barro Colorado Island. Each circle represents a species. Circles are colored and slightly jittered horizontally for clarity.

@@

\newpage

>>>>

Section  | Word Count  
------------- | -------------  
Summary | 197  
Main Text  | 6498  
Introduction | 828  
Methods | 3283  
Results | 883  
Discussion | 1504  
Keywords | 6

Item | Count  
------------- | -------------  
Figures in color | 9  
Tables | 2  
Supplementary Material |  1 file

@@

To obtain the daily dynamics of soil water potential in the whole-soil column, we locally parameterized and calibrated a 1-D hydrological water balance model. We then developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig. \@ref(fig:schematic)). We define ERD as the depth at which soil water potential best explained species-specific growth dynamics. The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf vulnerability curves by using existing data for eight species, while developing trait-based proxies for the rest (based on data for a total of 21 species). We validated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as independent observations. We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets for six to eight species that overlapped with our assessment of ERD. We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity. Finally, we analysed species-specific large tree mortality dynamics over 35 years and estimated the time a species’ spent beyond its critical hydraulic threshold to test whether ERD explained risk of drought-induced mortality.

@@

We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up. Extreme hydrological droughts were identified by depth $z$ as days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.

@@

\begin{equation}
\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big(z\,\Big|\, \textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|z,h}| r_{s|z,h} \ge 0.71\Big)\Big)
(\#eq:erd)
\end{equation}
where $L$ is the likelihood of $z$ given the observed growth time series, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m. We removed finely resolved soil layers <0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m as sufficient resolution for Model validation.

@@

\
We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ with ($n$ = `r nrow(mrate.depth.mean)`) and without deciduous species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`). To assess whether deeper ERD species escaped the experience of significant hydrological drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were more negative than species $\mathrm{\Psi}$~crit~. $\mathrm{\Psi}$~crit~ was defined as $\mathrm{\Psi}$~50,leaf~.

@@

Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species mortality rates significantly decreased with deeper ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig. \@ref(fig:erd-mort-census-evg)). When deciduous species were included, ERD explained lower but still significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig. \@ref(fig:erd-mort-census)). Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig. \@ref(fig:swp-droughts). While our drought assessment did not extend to the model spin-up period of 1985-1990, for which ERD explained significant mortality, a prior study found that during this period BCI experienced three extreme deficit water years, following the 1983 El-Niño year [@Condit.2017]. Analysis of a species' time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days beyond critical hydraulic thresholds and increasingly so over 2000-2005 and 2005-2010 (Fig. \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)). For the census interval 1990-1995, species with the shallowest ERD spent a high propotion of time beyond $\mathrm{\Psi}$~crit~ (Fig. \@ref(fig:erd-time)), but during the dry season (; Fig. \@ref(fig:swp-droughts), and did not have increased mortality rates (Fig. \@ref(fig:erd-mort-census). Overall, Species with ERD deeper than 1.7 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths deeper than 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).

@@

\
Our result--of lower mortality in deep-rooted trees--contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a south-asian seasonally tropical dry forest had higher mortality in a rare, prolonged drought. The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019 and @Ivanov:2012). In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ for depths deeper than 1 m did not cross $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa) for all the dry seasons and droughts. Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)`mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI are major factors in the different mortality responses. At BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths), \@ref(fig:swp-droughts)), whereas in CT2018 they were not. The contrast between the two studies highlights the combined role of seasonal precipitation input and site-specific hydrology in modulating deep ERD mortality-risk.

@@

\
Figure \@ref(fig:hydro-calib): ELM-FATES calibration and evaluation. Observations (red lines and points) versus simulations (gray lines) from 100 best-fit ensemble-member runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; **a**) and for an average of three vertical TDR probes (0-0.15 m; **b**); monthly stream discharge (Discharge, **c**) and monthly evapotranspiration from eddy flux tower (ET, **d**). Red points in (**a**) show average, manual plot-wide observations of VWC with 95% CI (error bars). All observations (red lines and points) were in-sample, except for VWC data from the horizontal probes (red lines in **a**) which were out-of-sample. Values in inset are average RMSE across the 100 best-fit simulations. For panel **a**, values at the top are for manual, plot-wide VWC and those at the bottom are for VWC from TDR probes. VWC is in units of cm^3^ cm^-3^, while ET and Discharge are in mm month^-1^.

\
Figure \@ref(fig:swp-droughts): Occurrence of extreme hydrological droughts by census intervals (horizontal panels) at three representative depths ($z$): 0.1, 0.6 and 1 m (vertical panels). Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\mathrm{\Psi}_{\textrm{soil},z}$) for a given DOY and depth $z$ are shown (same across census intervals). $\mathrm{\Psi}_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if at least one day of that year $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile for that day of year (DOY).

\
Figure \@ref(fig:ab-fit): Parameters $A$ (panel **a**) & $B$ (panel **b**) for the leaf vulnerability curves that were fitted to observed data on $K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~ for `r length(sp.exp.param$sp)` species (Eq. \@ref(eq:vc-exp)) vs. those that were predicted from a set of models based on trait-proxies. (Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod)). Goodness-of-fit ($R$^2^) and significance levels are given in inset.

\
Figure \@ref(fig:erd-iso): Modeled effective rooting depth (ERD; mean $\pm$ 1SE; m) versus dry-season stable isotopic concentration $\mathrm{\delta}$^2^H~xylem~ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island. $\mathrm{\delta}$^2^H~xylem~ data are from @Meinzer:1999.

\
Figure \@ref(fig:sp-erd): Modeled effective rooting depth (ERD; mean $\pm$ 1SE in m) for `r  length(erd.sp)` large, canopy species of Barro Colorado Island. A species ERD is defined as the median, across 100 hydrological realisations, of soil layer depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth for each realisation. As soil layers in the hydrological model are discretely resolved into exponentially increasing depths, so is the ERD axis. Deeper depths are shown with darker blue colors.

\
Figure \@ref(fig:erd-hyd): Pairwise correlations among modeled effective rooting depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama). Spearman's $r$ with red shades for positive $r$ values and blue shades for negative values. $R$^2^ and significance levels for linear model fits are given in panel insets. See Fig. S\@ref(fig:erd-hyd-full) for the full correlation matrix.

\
Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE in % yr^-1^) versus modeled effective rooting depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015). $R$^2^ and significance levels for linear model fits are given in panel insets. Model fits (blue lines) with SE (gray area) are only shown for periods with significant fits at $\alpha$ = 0.1, identified by census-intervals with a superscript *.

\
Figure \@ref(fig:erd-time): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars). Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~50,leaf~, and where $z$ is the soil depth matching species ERD. Standard error of the mean are shown over each bar. Data includes all the `r length(erd.mrate.sp)` canopy tree species on Barro Colorado Island that were included in the mortality analyses. Trends were similar for evergreen speceis alone.

@@
